---
title: "Conditional & Combined Plots"
---

```{r, echo = FALSE}
diabetes <- readRDS(here::here("data", "diabetes.rds"))
titanic <- readRDS(here::here("data", "titanic.rds"))
```

We can convert most univariate visualizations that we create with Base R graphics into conditional plots by specifying a
grouping factor. We then get a single figure with group-specific versions of the univariate visualization.

### Boxplots

We can use conditional boxplots to compare the distribution of a variable across groups. To create this type of figure,
we employ the so-called *formula interface* when calling the `boxplot()` function. In the following code, we use the
`diabetes` dataset to create sex-specific boxplots of disease progression.

```{r}
boxplot(progress ~ sex, data = diabetes)
```

When we use the formula interface, we define the plot's structure through an R formula object that we specify as the
first argument to `boxplot()`. In this formula, `progress ~ sex1`, we specify the plotted variable on the left-hand side
of the formula and we put the grouping factor on the right-hand side of the formula. We also have to use the `data`
argument to specify a data frame where `boxplot()` can find the variables we're referencing in the formula.

```{webr-r}
#| autorun: true
#| context: setup

download.file("https://openstatprog.com/raw-data/titanic.rds", "titanic.rds")
download.file("https://openstatprog.com/raw-data/diabetes.rds", "diabetes.rds")

titanic <- readRDS("titanic.rds")
diabetes <- readRDS("diabetes.rds")
```

::: practice

## Practice

Use the `titanic` dataset to create a conditional boxplot to visualize the distributions of `age` within the groups
defined by the `survived` variable.

- What does this figure tell you about the ages of survivors (`survived = 'yes'`) vs. non-survivors (`survived = 'no'`)?

*NOTE:* The `titanic` dataset is already loaded into the working directory of this *webr* session.

:::: {.panel-tabset}
## {{< var tab_title.editor >}}

```{webr-r}

```

## {{< var tab_title.solution >}}
```{r}
boxplot(age ~ survived, data = titanic)
```

The median age does not seem to differ much between the two groups, but the survivors' distribution appears to be more
symmetric, and there are fewer old survivors. The non-survivors' distribution is positively skewed by the
disproportionately higher number of fatalities among older passengers.
::::

:::

### Bar Plots

Conditional bar plots visualize two dimensional contingency tables. So, these types of figure allow us to evaluate how
the distribution of one categorical variable differs across levels of some other variable.

To create a conditional bar plot, we fist create a two-way contingency table, and then we plot that table with the
`barplot()` function In the following code we use the `titanic` dataset to visualize the conditional distributions of
ticket classes for the two survival groups.

```{r}
# Cross-tabulate ticket class and survival status
tab <- table(titanic$class, titanic$survived)

# Visualize the table created above
barplot(tab)
```

The default bar plot created above suffers from two important problems.

1. The bars for each ticket class group are stacked, so the meaning of the values shown on the y-axis is ambiguous
1. We don't know which ticket group is represented by which color.

We can solve these problems by plotting the bars beside each other by setting `beside = TRUE` and drawing a legend by
setting `legend = TRUE`.

```{r}
# Create a better visualization of the table
barplot(tab, beside = TRUE, legend = TRUE)
```

Now, we have a useful visualization. From this figure, we can clearly see that the relative survival rates of passengers
traveling in third class were much worse than the survival rates of other passengers.

::: practice

Use the `titanic` dataset to create a conditional bar plot to visualize the distributions of survival status within the
gender groups defined by the `sex` variable.

- Draw the bars beside each other
- Add a legend to the figure
- Give the x-axis a good label
- Give the plot a good title

What does this figure tell you about relative rates of survival between male and female passengers on the Titanic?

:::: {.panel-tabset}
## {{< var tab_title.editor >}}

```{webr-r}

```

## {{< var tab_title.solution >}}
```{r}
tab <- table(titanic$sex, titanic$survived)
barplot(tab, 
        beside = TRUE, 
        legend = TRUE,
        xlab = "Survived",
        main = "Survival Status of Titanic Passengers by Gender")
```

The relative survival rates are much better for female passengers. The number of males in the `survived = 'no'` group is
much higher than the number of females, and the patterns is reversed in the `survived = 'yes'` group.
::::

:::

### Histograms

When generating histograms, we can't really create the same kind of conditional plots that we can with boxplots and bar
plots, but we can overlay different plot layers to add useful information to our figure.

For example, in the following code, we draw a kernel density plot on top of an existing histogram.

```{r}
# Create a basic histogram of blood glucose levels
hist(diabetes$glu, probability = TRUE)

# Calculate the kernel density estimate of blood glucose levels
d <- density(diabetes$glu)

# Overlay the kernel density curve on the histogram
lines(y = d$y, x = d$x, col = "red")
```

In this case, we cannot use the `plot()` function to create the kernel density layer since `plot()` would just replace
the existing histogram with a new figure. Rather, we use the `lines()` function to add the appropriate curve as a new
layer in the existing plot. Note that we're explicitly extracting the `x` and `y` vectors from the object returned by
`density()` to define the x and y coordinates of the line.

The Base R graphical suite offers many functions like `lines()` that add layers to existing plots. In the next example,
we use the `abline()` function to add two dashed lines to denote the boundaries of the region containing the central 90%
of observations.

```{r}
# Create a basic histogram of average blood pressure levels
hist(diabetes$bp)

# Calculate the boundaries of the central interval containing 90% of observations
ci <- quantile(diabetes$bp, prob = c(0.025, 0.975))

# Add dashed lines to denoting the upper and lower 5% tail area cutoffs
abline(v = ci, lty = 2, lwd = 2)
```

### Scatterplots

We can apply the same type of layering process to mark up scatterplots with additional information. In the following
example, we add the least squares best fit line to a scatterplot showing the relation between average blood pressure,
`bp`, and blood glucose levels, `glu`.

```{r}
# Create a scatterplot of blood pressure against blood glucose
plot(bp ~ glu, data = diabetes, pch = 19)

# Estimate the linear regression of blood pressure on blood glucose
lmOut <- lm(bp ~ glu, data = diabetes)

# Add the best fit line to the scatterplot
abline(reg = lmOut, lwd = 2, col = "blue")
```

Again, we use the `abline()` function to overlay a line onto the existing plot. In this case, however, we define the
line based on the estimated regression coefficients stored in the `lmOut` object that we created with the `lm()`
function.

::: practice

Use the `diabetes` dataset to create a scatterplot with average blood pressure, `bp`, on the y-axis and body mass index,
`bmi` on the x-axis.

- Draw the points as solid black squares.
- Add better axis labels.
- Add a better plot title.

Add a horizontal line to the scatterplot showing the mean of the `bp` variable.

- Draw the line as a dotted black line.

Add the best fit line from a linear regression model where `bmi` predicts `bp`.

- Draw the line as a solid, red line.
- Increase the width of the line to improve legibility.

What does this figure tell you about relation between body mass index and average blood pressure?

*NOTE:* The `diabetes` dataset is already loaded into the working directory of the *webr* session.

:::: {.panel-tabset}
## {{< var tab_title.editor >}}

```{webr-r}

```

## {{< var tab_title.hint >}}

To implement the visual adjustments to the points and lines, you will need to adjust the values of the following
parameters in various function calls.

- `pch`
- `lty`
- `lwd`
- `col`

## {{< var tab_title.solution >}}
```{r}
# Create the scatterplot
plot(bp ~ bmi,
     data = diabetes,
     pch = 15,
     ylab = "Blood Pressure",
     xlab = "Body Mass Index",
     main = "Association between BMI and Average Blood Pressure")

# Add the horizontal rule
abline(h = mean(diabetes$bp), lty = 3)

# Estimate the regression model
lmOut <- lm(bp ~ bmi, data = diabetes)

# Add the best fit line from the above regression
abline(reg = lmOut, lwd = 2, col = "red")
```

We see what looks like a strong, positive effect of BMI on blood pressure. In other words, higher BMI tends to be
associated with higher blood pressure levels.
::::

:::
