---
title: "Scatterplots"
---

Scatterplots are one of the most fundamental tools for visualizing the relationship between two continuous variables (or
variables that we wish to treat as continuous). Each point on the plot represents one observational unit's values on the
two plotted variables. This makes scatterplots particularly useful for identifying trends, clusters, and extreme values
in the data.

```{webr-r}
#| autorun: true
#| context: setup

download.file("https://openstatprog.com/raw-data/diabetes.rds", "diabetes.rds")
diabetes <- readRDS("diabetes.rds")
```

```{r, echo = FALSE, include = FALSE}
diabetes <- readRDS(here::here("data", "diabetes.rds"))
```

We make a simple scatterplot with the `plot()` function. For example, the following code uses the `diabetes` dataset to
visualize the bivariate association between Body Mass Index, `bmi`, and Total Cholesterol, `tc`.

```{r}
plot(y = diabetes$tc, x = diabetes$bmi)
```

In this example, every point represents a different patient's value of BMI and their level of total cholesterol. The
points form a cloud with a slight upward trend, suggesting that higher BMI values tend to be associated with higher
cholesterol levels, although the variability is substantial.

We can improve the readability by adding axis labels, a title, and adjusting the axis limits. To do so we adjust the
following parameters.

- `ylab`, `xlab`: Define new axis labels
- `main`: Add a plot title
- `ylim`, `xlim`: Define the ranges of the y-axis and x-axis

```{r}
plot(y = diabetes$tc,
     x = diabetes$bmi,
     ylab = "Total Cholesterol",
     xlab = "Body Mass Index",
     main = "Relation between BMI and Cholesterol",
     ylim = c(0, 350),
     xlim = c(0, 50))
```

When we apply `plot()` to a data frame containing only numeric columns, we generate a scatterplot matrix: a grid of
scatterplots showing all the pairwise relationships.

```{r}
plot(diabetes[c("age", "bmi", "tc", "glu", "progress")])
```

::: practice

## Practice

Using the `diabetes` dataset, make a scatterplot of *total cholesterol* on the y-axis against *age* on the x-axis. Add
meaningful axis labels and a title.

- The `diabetes` dataset is already loaded in the working directory of this *webr* session.

:::: {.panel-tabset}
## {{< var tab_title.editor >}}

```{webr-r}

```

## {{< var tab_title.solution >}}

```{r}
plot(y = diabetes$tc,
     x = diabetes$age,
     ylab = "Total Cholesterol",
     xlab = "Age",
     main = "Relation between Age and Cholesterol")
```
::::

:::
