---
title: "Module Template"
page-navigation: false
format:
  html: 
    toc: false
---

```{r, echo = FALSE, include = FALSE}
source(here::here("R", "buttons.R"))
```

Welcome to the module on basic data visualization in R. Graphs are a crucial tool in data analysis because they allow us to explore, summarize, and communicate patterns in data that would be difficult to detect through numbers alone. Well-chosen visualizations can reveal trends, clusters, correlations, and outliers at a glance, helping to guide further analysis and support clear, persuasive communication of results.

In this module, we will explore the main plotting capabilities built into base R, focusing on four foundational chart types: scatterplots for examining relationships between two numeric variables, histograms for understanding the distribution of a single variable, boxplots for summarizing distributions and comparing groups, and density plots for visualizing the smooth distribution of continuous data. 

Each section combines conceptual explanation with hands-on examples to show how to create and customize these plots in R, highlighting how they can be adapted to different datasets and analysis goals.

## Learning Goals

After completing this module, you will be able to:

1. Create scatterplots to investigate relationships between two variables.
2. Build histograms to examine the distribution of a numeric variable.
3. Use boxplots to summarize distributions, detect outliers, and compare groups effectively.
4. Produce density plots to visualize approximations of a variable’s distribution.

Click the button below to get started with the first tutorial.

```{r}
#| echo: false
#| results: asis
begin_button("/tutorials/basic_graph/index.html")
```
