---
title: "Bar Plots"
---

```{r, echo = FALSE}
titanic <- readRDS(here::here("data", "titanic.rds"))
```

Bar plots are the analogue of histograms for discrete variables. Whereas the columns in a histogram are defined by
counting the number of cases that fall into arbitrary bins, the bars in a bar plots represent the number of cases that
belong to each level of a categorical variable.

We use the `barplot()` function to create bar plots in Base R. As with kernel density plots, however, we first need to
summarize the data before creating the visualization. Specifically, we use the `table()` function to create a frequency
table, and we then plot that table with `barplot()`. The following code visualizes the distribution of ticket classes in
the `titanic` dataset.

```{r}
# Tabulate the counts of each ticket class
classTable <- table(titanic$class)

# Create a bar plot to visualize the frequency of ticket classes
barplot(classTable)
```

To plot the counts as proportions, we convert the data in the table before plotting.

```{r}
# Convert counts to proportions
classProps <- classTable / sum(classTable)

# Create a bar plot to visualize the frequency of ticket classes
barplot(classProps)
```

As with all of our Base R plotting, we can tweak the appearance in many ways.

```{r}
barplot(classTable,
        horiz = TRUE,
        main = "Classes of Titanic Tickets",
        ylab = "Ticket Class",
        xlab = "Number of Passengers",
        col = rgb(181, 137, 0, max = 255),
        border = rgb(0, 43, 54, max = 255))
```
