---
title: "Boxplots"
---

```{r, echo = FALSE}
diabetes <- readRDS(here::here("data", "diabetes.rds"))
titanic <- readRDS(here::here("data", "titanic.rds"))
```

The boxplot (AKA, box-and-whiskers plot) is one of the most enduring contributions of Tukey's seminal book [*Exploratory
Data Analysis*][eda_book]. A boxplot is a visualization of a numeric variable's so-call "five-number-summary" that
quantifies a variable's distribution through five order statistics: minimum, first quartile, median, third quartile,
maximum. Boxplots are particularly useful for evaluating the spread and symmetry of a distribution and checking for
extreme values.

We use the `boxplot()` function to create a boxplot with Base R graphics. For example, the following code creates a
basic boxplot of the `progress` variable from the `diabetes` dataset.

```{r}
boxplot(diabetes$bmi)
```

In this case, the plot indicates some positive skew: the median falls closer to the first quartile than the third
quartile, and the upper whisker is longer than the lower whisker.

### Defining the Whiskers

The "box" part of a box-and-whiskers plot is always defined by first and third quartiles, $Q_1$ and $Q_3$, with the
median drawn somewhere inside the box. So, this box tells us something about the symmetry and spread of the middle 50%
of the distribution.

The "whiskers", on the other hand, tell us about the tails of the distribution. We can defined the whiskers in several
different ways to target specific information. In the default boxplot, the whiskers are defined in terms of the
*inner-quartile range* $\textrm{IQR} = Q_3 - Q_1$.

- Lower whisker: $Q_1 - 1.5 \times \textrm{IQR}$
- Upper whisker: $Q_3 + 1.5 \times \textrm{IQR}$

We can change the weight applied to the IQR via the `range` argument. The following code defines the whisker margin as
$2 \times \textrm{IQR}$.

```{r}
boxplot(diabetes$bmi,
        range = 2,
        xlab = "Body Mass Index",
        main = "Margin = 3 * IQR")
```

To use the minimum and maximum values to define the whiskers, we set `range = 0`.

```{r}
boxplot(diabetes$bmi,
        range = 0,
        xlab = "Body Mass Index",
        main = "Min/Max Whiskers")

```

### Boxplots for Outlier Analysis

Any observations that fall outside the interval defined by the whiskers are plotted as individual points. This behavior
makes boxplots a handy way to check for extreme values (i.e., outliers). If we set the whiskers so they define the range
of plausible values, any points that fall outside the whiskers are possible outliers. 

A common convention is to use a margin of $1.5 \times \textrm{IQR}$ to flag "possible" outliers and a margin of $3
\times \textrm{IQR}$ to flag "probable" outliers.

```{r, echo = FALSE}
nOut1.5 <- boxplot.stats(diabetes$bmi, coef = 1.5)$out |> length()
nOut3   <- boxplot.stats(diabetes$bmi, coef = 3)$out |> length()
```

```{r}
boxplot(diabetes$bmi,
        range = 1.5,
        xlab = "Body Mass Index",
        main = "Possible Outliers")
```

```{r}
boxplot(diabetes$bmi,
        range = 3,
        xlab = "Body Mass Index",
        main = "Probable Outliers")
```

In this case, we see that `r nOut1.5` observations are flagged as possible outliers, and `r nOut3` observations are
flagged as probable outliers.

```{webr-r}
#| autorun: true
#| context: setup

download.file("https://openstatprog.com/raw-data/titanic.rds", "titanic.rds")
titanic <- readRDS("titanic.rds")
```

::: practice

## Practice

Use the `titanic` data to create a boxplot of `age`.

- Use a margin of $2 \times \textrm{IQR}$ to define the whiskers.
- Add a better plot title
- Add a better label for the x-axis

*NOTE:* The `titanic` dataset is already loaded in the working directory of this *webr* session.

:::: {.panel-tabset}
## {{< var tab_title.editor >}}

```{webr-r}

```

## {{< var tab_title.solution >}}

```{r}
boxplot(titanic$age,
        range = 2,
        xlab = "Age",
        main = "Ages of Titanic Passengers")
```
::::

:::

[eda_book]: https://www.google.com/books/edition/Exploratory_Data_Analysis/UT9dAAAAIAAJ?hl=en
