---
title: "Histograms"
---

```{r, echo = FALSE}
diabetes <- readRDS(here::here("data", "diabetes.rds"))
```

Histograms are one of the most straightforward and useful ways to visualize the distribution of a numeric variable. A
histogram shows the distribution of a single numeric variable by dividing its range into bins (intervals), counting how
many observations fall into each bin, and visualizing these counts as bars in a plot. Histograms are meant to represent
a variable's probability distribution, so they're useful for understanding the shape, spread, and skewness of numeric
variables

The `hist()` function in Base R creates histograms. For example, the following code visualizes the distribution of blood
glucose levels in the diabetes dataset.

```{r}
hist(diabetes$glu)
```

### Adjusting the Basic Plot

By default, `hist()` will try to best represent the distribution by applying the [Sturges Formula][sturges] to define
the bin width We can alter this behavior in several ways through the `breaks` argument.

- Specify a different built-in algorithm
- Provide a vector of cut points giving the bin margins
- Request a fixed number of bins
- Define our own function to create the bins

```{r}
hist(diabetes$glu,
     main = "Define bins via Scott's formula",
     breaks = "scott")
hist(diabetes$glu,
     main = "Define bins with a vector",
     breaks = seq(50, 130, 10))
hist(diabetes$glu, main = "Ask for 5 bins", breaks = 5)
hist(diabetes$glu, main = "Ask for 50 bins", breaks = 50)
```

The `hist()` function accepts many parameters to customize the plot's appearance. For example, in the next code chunk,
we adjust the following parameters.

- `main`: Add a better title
- `col`: Change the fill color of the bars
- `border`: Change the border color of the bars
- `probability`: Switch the y-axis to a probability scale

```{r}
hist(diabetes$glu,
     main = "Colorful Histogram",
     col = "pink",
     border = "blue",
     probability = TRUE)
```

Setting `probability = TRUE` converts the y-axis to a probability scale and makes the histogram’s total area sum to 1.
Hence the bar heights represent the proportion of the sample that falls into each bin, rather than a straight count of
the observations in each bin.

### Graphical Parameters

::: {.column-margin}

:::: resource

The tutorial linked below gives a nice overview of the various graphical parameters you can use to adjust the appearance
of Base R visualizations.

[STHDA Tutorial: R Graphical Parameters](https://www.sthda.com/english/wiki/graphical-parameters)

::::

:::

The adjustments you see above are only a tiny drop in a vast bucket of possibilities. We can tweak nearly every aspect
of our visualization's appearance by adjusting the values of the built-in *graphical parameters*. There are many such
parameters. To see a full list of the graphical parameters, read the documentation for the `par()` function.

```{r, eval = FALSE}
?par
```

```{webr-r}
#| autorun: true
#| context: setup

download.file("https://openstatprog.com/raw-data/diabetes.rds", "diabetes.rds")
diabetes <- readRDS("diabetes.rds")
```

::: practice

## Practice

Make an histogram showing the age distribution in the `diabetes` dataset.

- Adjust the bin width to create a fine-grained visualization with many, small bins
- Set the fill-color of the columns to green
- Add a better label for the x-axis
- Add a better plot title

*NOTE:* The `diabetes` dataset is already loaded in the working directory of this *webr* session.

:::: {.panel-tabset}
## {{< var tab_title.editor >}}

```{webr-r}

```

## {{< var tab_title.solution >}}

```{r}
hist(diabetes$age,
     col = "green",
     breaks = 50,
     xlab = "Age",
     main = "Age Distribution")
```
::::

:::

### Kernel Density Plots

A kernel density plot provides a smoothed estimate of a continuous variable’s distribution. You can think about a kernel
density plot as a smoothed version of a histogram. Essentially, we create a density plot by drawing a line that connects
the tops of the columns from a special type of fine-grained histogram into a single continuous curve.

In Base R, we don't have a single function to create a kernel density plot. We first compute a density estimate with the
`density()` function, and then we plot the estimated density object. For example, the following code visualizes the
distribution of BMI with a kernel density plot.

```{r}
d <- density(diabetes$bmi)
plot(d)
```

As with any other Base R visualization, we can tweak the appearance of a kernel density plot through graphical
parameters.

```{r}
plot(d, main = "Estimated Density of BMI", col = "blue")
```

[sturges]: https://doi.org/10.1080/01621459.1926.10502161
