---
title: "Boxplots"
---

```{r, echo=FALSE}
diabetes <- readRDS("data/diabetes.rds")
titanic <- readRDS("data/titanic.rds")
```

A boxplot summarizes a numeric variable’s distribution through five key statistics: the minimum, first quartile, median, third quartile, and maximum. It also highlights potential outliers. Boxplots are particularly useful for comparing distributions across groups and are a compact alternative to histograms when you want to quickly compare medians, spreads, and detect asymmetry or extreme values without showing the full shape of the distribution.

A basic boxplot of the `progress` variable looks like this:

```{r}
boxplot(diabetes$progress)
```

In this case, the plot shows a median roughly in the middle of the box, fairly balanced whiskers, and no evident extreme outliers, suggesting a relatively symmetric distribution. In boxplots, outliers are typically displayed as individual points beyond the whiskers, which extend to 1.5 times the interquartile range from the box.

Boxplots in base R accept many of the same parameters as the plot() function, such as axis labels, main titles, and limits, allowing for consistent styling across different plot types. We can also rotate the boxplot horizontally and change the whisker range:

```{r}
boxplot(diabetes$progress,
        horizontal = TRUE,
        range = 3,
        xlab = "Disease Progression")
```

To compare groups, we use a formula, where the variable of interest is plotted against a factor:

```{r}
boxplot(progress ~ sex, data = diabetes, col = "violet")
```

This produces side-by-side boxplots for each group in `sex`, allowing a quick visual comparison of tendencies and spread.

```{webr-r}
#| autorun: true
#| context: setup
#| 
download.file("https://raw.githubusercontent.com/p3piit/exp_rawdata/main/diabetes.rds", "titanic.rds")
titanic <- readRDS("titanic.rds")
library(magrittr)
```


::: practice

## Practice

Use base R graphics and the `titanic` data to create conditional boxplots, where plots of `age` are conditioned on `survived`. What does this figure tell you about the ages of survivors (`survived` = 1) vs. non-survivors (`survived` = 0)?

:::: {.panel-tabset}
## {{< var tab_title.editor >}}

```{webr-r}

```

## {{< var tab_title.solution >}}

As we can see, the age distributions of survivors and non-survivors are quite similar. Both groups have nearly the same median age and a comparable spread of ages, indicating that age alone does not appear to be a strong distinguishing factor between survival and non-survival in this dataset. 

```{r}
boxplot(age ~ survived, data = titanic)
```
::::

:::














